---
title: "Project 2"
author: "James Spalding, Ryan Winder"
date: "2024-04-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, include = F)

#Libraries
library(tidyverse)
library(readxl)
library(DescTools)
library(gt)
library(gtExtras)
library(gridExtra)
library(gtable)
library(ggpubr)
library(boot)
```

```{r}
data = read.csv("CervicalCancer.csv") %>% select(-"STDs.AIDS") #None of the patients had AIDS

#split to positive and negative
positiveData = data %>%
  filter(Biopsy == 1)

negativeData = data %>%
  filter(Biopsy == 0)
```

## Introduction

Cervical cancer is a disease that affects around 200,000 women per year in the United States. It has minor symptoms and is difficult to detect without a screening. However, according to *American Cancer Society*, if detected early, cervical cancer is one of the most successfully treatable cancers with over an 80% survival rate if detected in its first year. 

For this reason, the main goal of our study is to determine which lifestyle and sexual factors significantly contribute to the presence of cervical cancer in patients so they can get timely screenings and treatment.

## Exploring the Data

```{r, include = T, fig.align = 'center'}
##### Summary Stats Table #####

Term       = c("Age",
               "Number of sexual partners",
               "Age of first sexual intercourse", 
               "Number of Pregnancies",
               "Smoke",
               "Years of Smoking",
               "Packs Smoked/Year",
               "Hormonal Contraceptives", 
               "Years used Hormonal Contraceptives",
               "IUD",
               "Years used IUD",
               "Has had STD",
               "Condylomatosis", 
               "Syphilis",
               "Pelvic Inflammatory Disease",
               "Genital Herpes",
               "Molluscum Contagiosum",
               "HIV",
               "Hepatitis B",
               "HPV",
               "Positive for Cervical Cancer")

Ratio = c()
Count = c()
Mean = c()
Min = c()
Max = c()
SD = c()

#Obtain descriptive stats
for(i in names(data)){
  #Determine if percent or mean is applicable
  if (length(unique(data[[i]])) == 2){
    Ratio = c(Ratio, paste0(round(mean(data[[i]]),4) * 100, "%"))
    Count = c(Count, length(which(data[[i]] == 1)))
    Mean  = c(Mean, "")
    Min   = c(Min, "")
    Max   = c(Max, "")
    SD    = c(SD, "")
  }else{
    Ratio = c(Ratio, "")
    Count = c(Count, "")
    Mean  = c(Mean, round(mean(data[[i]]),2))
    Min   = c(Min, min(data[[i]]))
    Max   = c(Max, max(data[[i]]))
    SD    = c(SD, round(sd(data[[i]]),2))
  }
}

#Decided to split into 2 types for better presentation...

#Binary variables
binTerm = Term[c(5,8,10,12:21)]
Percentage = Ratio[c(5,8,10,12:21)]
Count = Count[c(5,8,10,12:21)]

#Quantitative variables
quantTerm = c(Term[c(1:4, 6,7,9)], rep("",6))
Mean = c(Mean[c(1:4, 6,7,9)], rep("",6))
Min  = c(Min[c(1:4, 6,7,9)], rep("",6))
Max  = c(Max[c(1:4, 6,7,9)], rep("",6))
SD   = c(SD[c(1:4, 6,7,9)], rep("",6))


records = as.data.frame(t(as.data.frame(rbind(binTerm, Count, Percentage,
                                              quantTerm, Mean, Min, Max, SD))))

records %>%
  gt() %>%
  gt_theme_nytimes() %>%
  tab_header(title="Table 1: Variables explained.") %>%
  tab_spanner(label="Binary Terms", columns = c(binTerm, Count, Percentage)) %>%
  tab_spanner(label="Quantitative Terms", columns = c(quantTerm, Mean, Min, Max, SD)) %>%
  cols_label(binTerm = " ", quantTerm = " ")
```

In our study, there are a total of 647 patients; 24 are positive for cervical cancer, or around 4%. A few noteworthy statistics in *table 1* are that the mean age of patients is around 27 years old, around 65% have used hormonal contraceptives for a mean of 2.25 years, and around 9% have had some sort of STD with HIV being the most common at an overall percentage of 1.85%. Using this information, we are able to perform various tests to determine which of these variables are significant to the presence of cervical cancer.

## Statistical Methods

*All analysis conducted in R 4.2.2*

```{r, eval = F}
#correlation test

corPerm = function(x, y, nperm=10000){
  r.perm=rep(0,nperm)
  r.obs=cor(x,y)
  
  for(i in 1:nperm){
    y.permuted=sample(y)
    r.perm[i]=cor(x,y.permuted)
  }
  
  return(sum(r.perm>=r.obs)/nperm)
}


corTable = data.frame(var1 = c(),
                      var2 = c(),
                      pvals = c())

for(p in 1:21){
  print(names(data)[p])
  for(q in 1:21){
    print(names(data)[q])
    temp = data.frame(names(data)[p],
                      names(data)[q],
                      corPerm(data[[p]], data[[q]], nperm = 1000))
    corTable = rbind(corTable, temp)
  }
}

corTable2 = corTable[2:441,]
corTable2 = corTable2[-seq(from = 22, to = 440, by=22),] #remove cor with itself

sig = corTable2[which(corTable2$corPerm.data..p....data..q....nperm...1000. > .7),]
```

To begin, we decided to look into correlation between variables. As the data is non-parametric, we will use Pearson's $\rho$ coefficient and permute the data 1,000 times per test to save time. With 21 variables, there are 441 different correlation values; after removing correlation with itself, there are 420 values. Of these, around 100 had a $\rho$ value greater than 0.7. When looking specifically at the biopsy variable, the following variables have a high correlation: HPV, Hepatitis, Pelvic Inflammatory Disease, Molluscum Contagiosum, Syphilis, and Number of Sexual Partners. All of these variables are significant, and many will appear later in other tests.

Next, we are interested in looking at the quantitative variables in our data set to uncover if any are significant in detecting the presence of cervical cancer in patients. For this, we are going to use a two-sided two-sample permutation test with the first sample being patients with cervical cancer and the other being patients without. We are interested in looking to uncover significant quantitative variables in terms of the mean, variance, and medians. For our tests we are using a significance level($\alpha$) equal to .05.

```{r, warning=FALSE, include=T}
attach(data)
difference_function <- function(x, test_to_run = mean) {
  function_list <- list(
    Cancer = c(x[data$Biopsy == 1]),
    No_Cancer = c(x[data$Biopsy == 0])
  )
  
  
  dat1 <- function_list$Cancer
  dat2 <- function_list$No_Cancer
  
  m=length(function_list$Cancer)
  n=length(function_list$No_Cancer)
  tot=m+n
  Dobs=test_to_run(dat1)-test_to_run(dat2)

  nperm=10000
  Dperm=rep(0,nperm)
  alldat=c(dat1,dat2)

  for(i in 1:nperm){
    index=sample(1:tot,m)
    newdat1=alldat[index]
    newdat2=alldat[-index]
    Dperm[i]=test_to_run(newdat1)-test_to_run(newdat2)
  }

  pvalue=sum(abs(Dperm)>=abs(Dobs))/nperm
  pvalue
}

p_values_mean <- c()
p_values_var <- c()
p_values_median <- c()

vars <- c("Age","Number.of.sexual.partners", "First.sexual.intercourse", "Num.of.pregnancies",
          "Smokes..years.", "Smokes..packs.year.", "Hormonal.Contraceptives..years.",
          "IUD..years.")

for(i in 1:length(vars)) {
  p_values_mean[i] <- difference_function(data[,vars[i]])
  p_values_var[i] <- difference_function(data[,vars[i]], var)
  p_values_median[i] <- difference_function(data[,vars[i]], median)
}

explanatory_variables <- c("Age","Number of sexual partners","Age of first sexual intercourse","Number of pregnancies",
                           "Years smoking", "Packs smoked a year", "Years using Hormonal Contraceptives",
                           "Years using IUD")

permutation_tests <- data.frame(
  Variable = explanatory_variables,
  Pval_mean = p_values_mean,
  Pval_var = p_values_var,
  Pval_median = p_values_median
)

permutation_tests %>%
  gt() %>%
  gt_theme_nytimes() %>%
  tab_header(title="Table 2: Probabilities of Two-Sample Permutation Tests")
```

As you can see above in *table 2*, there was no significant quantitative variable found in terms of the means or median between the two samples. Then for the two-sided two-sample permutation test for a difference in the variance, a significant difference was found (with a p-value of around .02) for the variable years using hormonal contraceptives. Also, this variable gave the lowest p-value out of all the quantitative variables when testing for a difference in sample means, giving a p-value of around .09.

Next, we are interested in testing all of the qualitative variables in our data set to uncover significant differences between patients diagnosed with cervical cancer to those without. For this, we are going to use Fisher's Exact Test for 2x2 tables. We are able to do this by grouping each categorical variable with the categorical variable Biopsy, which is the variable informing us whether a patient is diagnosed with cervical cancer or not. We will be using a significance level($\alpha$) of .05 for this test.

```{r, include = T, message=F}
# Fisher's Exact Test for 2x2

fisher = function(x) {
  
  test = matrix(c(sum(data[,x]==1 & data[,"Biopsy"]==1),
         sum(data[,x]==0 & data[,"Biopsy"]==1),
         sum(data[,x]==1 & data[,"Biopsy"]==0),
         sum(data[,x]==0 & data[,"Biopsy"]==0)), nrow=2)
  
  fisher.test(test, alternative="greater")$p.value
}

vars2 = c("Smokes","Hormonal.Contraceptives","IUD","STDs","STDs.condylomatosis",
           "STDs.syphilis","STDs.pelvic.inflammatory.disease","STDs.genital.herpes",
           "STDs.molluscum.contagiosum","STDs.HIV","STDs.Hepatitis.B","STDs.HPV")

p_values = c()

for(j in 1:length(vars2)) {
  p_values[j] = round(fisher(vars2[j]),2)
}

explanatory_variables2 = c("Smokes", "Hormonal Contraceptives", "IUD",
                            "STDs", "Condylomatosis", "Syphilis", "Pelvic Disease",
                            "Genital Herpes", "Molluscum Contagiosum", "HIV",
                            "Hepatitis B", "HPV")

permutation_tests2 = data.frame(
  Variable = explanatory_variables2,
  Pval = p_values
)

permutation_tests2 %>%
  gt() %>%
  gt_theme_nytimes() %>%
  tab_header(title="Table 3: Probabilities of Fisher's Exact Tests")
```

```{r, include = T, fig.align = 'center'}
#print("HPV, pval = 1")
m1 = data.frame(matrix(c(sum(data[,"STDs.HPV"]==1 & data[,"Biopsy"]==1),
         sum(data[,"STDs.HPV"]==0 & data[,"Biopsy"]==1),
         sum(data[,"STDs.HPV"]==1 & data[,"Biopsy"]==0),
         sum(data[,"STDs.HPV"]==0 & data[,"Biopsy"]==0)), nrow=2))

#print("genital herpes, pval < .05")
m2 = data.frame(matrix(c(sum(data[,"STDs.genital.herpes"]==1 & data[,"Biopsy"]==1),
         sum(data[,"STDs.genital.herpes"]==0 & data[,"Biopsy"]==1),
         sum(data[,"STDs.genital.herpes"]==1 & data[,"Biopsy"]==0),
         sum(data[,"STDs.genital.herpes"]==0 & data[,"Biopsy"]==0)), nrow=2))

#print("stds, pval < .05")
m3 = data.frame(matrix(c(sum(data[,"STDs"]==1 & data[,"Biopsy"]==1),
         sum(data[,"STDs"]==0 & data[,"Biopsy"]==1),
         sum(data[,"STDs"]==1 & data[,"Biopsy"]==0),
         sum(data[,"STDs"]==0 & data[,"Biopsy"]==0)), nrow=2))


grid.arrange(
  tableGrob(m1, theme = ttheme_minimal(), rows = c("",""), cols = c("HPV:","P=1")),
  tableGrob(m2, theme = ttheme_minimal(), rows = c("",""), cols = c("Herpes:","P<.05")),
  tableGrob(m3, theme = ttheme_minimal(), rows = c("",""), cols = c("STDs:","P<.05")),
  nrow=1)


```


As you can see above in *table 3*, there was found to be three variables with significant differences between patients with cervical cancer to those without. Those include the variable STDs, which returned a p-value of .019, the variable Condylomatosis, which returned a p-value of .032, and lastly the variable Genital Herpes, which returned a p-value of .037. The variable HIV is noticeable close with a p-value of .069.

```{r}
#bootstrap for multi reg
obs.mod = lm(Biopsy~., data = data)


e = obs.mod$residuals
modsummary = summary(obs.mod)

fObs = modsummary$fstatistic[1]
tObs = c()
for(i in 2:21){
  tObs = c(tObs, modsummary$coefficients[i,3])
}

bootData = data %>%
  select(-Biopsy) ; bootData$e = rep(999, 647)

nboot=1000
tBoot = array(dim = c(20,1,nboot))
fBoot = c()
for (i in 1:nboot){
  eboot = sample(e,replace=TRUE)
  bootData$e = eboot
  
  boot.mod = lm(e~., data=bootData)
  modsummaryP=summary(boot.mod)

  for(j in 2:21){
    tBoot[j-1,1,i] = modsummary$coefficients[j,3]
  }
  
  fBoot = c(fBoot,modsummaryP$fstatistic[1])
}

#Overall test
print("At least 1 is significant:")
sum(fBoot>=fObs)/nboot #.03 that at least 1 is significant

#Individual tests
for(i in 1:20){
  if(sum(abs(tBoot[i,,])>=abs(tObs))/nboot < .06){
    print(names(data)[i])
    print(sum(abs(tBoot[i,,])>=abs(tObs))/nboot)
  }
}

#Only STDs.molluscum.contagiosum appears significant. Maybe dont include this.
```

Finally, we wanted to create a regression model to determine which factors are useful in predicting the presence of cervical cancer in a patient. Our null hypothesis is that all factors have a $\beta$ value of 0, and thus are unhelpful in predicting the biopsy value. The alternative is that at least 1 of the factors has a $\beta$ value not equal to 0. By using a bootstrap test with 10,000 iterations, we were able to obtain a p-value for this test, which came out as 0.039, and thus we reject the null hypothesis.

Next, we performed a bootstrap test, again with 10,000 iterations, on each of the 20 variables. Doing this, we were unfortunately only able to find 1 variable with a significant $\beta$ value. The variable was Molluscum Contagiosum, and it had a p-value of 0.05.



## Results and Conclusion



*Describe results from tests*

*Explain what the results mean/conclude*